---
title: "Social Determinants of Wage Gaps"
author: "Claire Lai, Trang Tran, Tony Zhang, Jiali Liu"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

|Element | Brief Description|
|---------------| --------------
|Code name | Dataforjustice |
|Project title| Gender Wage Gap and The Intersection with Racial Identity |
|Authors | Trang T Tran (trangtt@uw.edu); Claire Lai (cglai@uw.edu); Jiali Liu (jiali123@uw.edu); Tony Zhang (yzhangss@uw.edu)|
|Affiliation |  INFO-201: Technical Foundations of Informatics - The Information School - University of Washington |
|Date | Winter 2022|
|Abstract | Our main concern is how intersectionality contribute to shaping the system of oppression based on social identity. Specifically, our question is about how racial identity intertwined with gender to affect the current wage gap. To address this question, we will analyze the difference between income of individuals at the same educational level with different gender and racial identity. |
|Keywords | social justice, gender, race, wage gap |
|1.0 Introduction | In this project, we will examine the additional effect of race on the existing gender wage gap, in an effort of highlighting the role of intersectionality when investigating systemic oppression. It is important, as justice will not be served if during our process of fighting against injustices, we overlook the struggles of some. Those are people who take on more than one marginalized identities. Specifically, in this project, we are focusing on women of color. Advocacy for a just system will benefit not one but everyone, as it is a "co-liberation" process. Through this project, we hope to break it down to better understand how each minoritized identity factors in the systemic injustices, regarding income. As such, we hope to shed light on the generational wealth gap in America.|
|2.0 Design Situation | The topic that this project will focus on is the race and gender wage gap, research into the problem spaced showed that this is an issue across the US. One of the sources, ["Race and Gender Wage Gaps in the Market for Recent College Graduates"](https://alliance-primo.hosted.exlibrisgroup.com/permalink/f/lvbsh/TN_cdi_proquest_miscellaneous_61550254) shows that after controlling for education level and college major a study showed that white male graduates earned 10 to 15 percent more when compared to other graduates of different races and genders (Weinberger). The issues of the wage gap is global however, as even a cursory glance at the data set, which was sourced in Russia, shows that there are differences. </p> The source of the data contributes to **project frame** which will focus on and help tackle the wage gap inequality between races and genders in Russia due to the origin of the data set. </p> The **human values** behind this project are ones of equality across genders and races. One of the ways to fight against inequality is to show where the inequalities are and that is what this project aims to do.</p>  The **direct stakeholders** of this project are the men and women that are affected by the wage gap due to their race and gender. Figure 1 from the source [“The Gender Wage Gap in the United States and Cross Nationally”](https://alliance-primo.hosted.exlibrisgroup.com/permalink/f/lvbsh/TN_cdi_proquest_journals_1616822208) shows the median distribution of wages and Blacks and Latinos/Latinas are just a few of the direct stakeholders that are negatively affected (Misra, Murray-Close). </p> The **indirect stakeholders** in this situation are those who are not in the working world at the moment but plan to join it. Even though they are not directly impacted by the current wage gap, once they start to work they will be and so even as ‘outsiders’ they still have a stake in the issue. </p> While exposing the wage gap between genders and races has the **benefit** of bringing to light the inequalities of the working world, a possible **harm** arises through the more mainstream genders of "male" and "female." The terms "male" and "female" are used in all three sources including the source [“Ceteris non paribus: The intersectionality of gender, race, and region in the gender wage gap”](https://alliance-primo.hosted.exlibrisgroup.com/permalink/f/lvbsh/TN_cdi_gale_infotracacademiconefile_A523188234), and are also used in the data set, which erases the identities of those who are gender-fluid, nonbinary, and the like and fails to recognize how identifying as a gender that is outside of the "mainstream" affects the wage inequality faced in the working world. |
|3.0 Research questions | The three research questions are **How does racial identity affect income outcomes?**, **How does gender identity affect income outcomes?** and **How does the intersectionality of racial and gender identities affect income outcomes?** In this project these questions will be grouped by education level to control for wage difference between educational levels, as it is one of the closest proxies for income level. </p> The preliminary research done in part 2 clearly shows there is a wage gap between men and women as well as between races. However, while some may acknowledge that this is an issue they question the magnitude of it, their perception on who it affects and how it affects certain groups are skewed. They believe that the issue is not a systematic one but instead lies on the jobs people of different races and genders choose. These research questions are designed to investigate the magnitude of impacts that race and gender have on income levels, as well as how the intersection between them might impact educational outcomes differently. |
|4.0 The Dataset | The dataset we used comes from [__Kaggle__](https://www.kaggle.com/ljanjughazyan/wages), an online platform where scientists and machine learning practitioners can share datasets. A Kaggle user uploaded the dataset four years ago and published it on Kaggle and made it accessible to the public. In making its dataset publicly available, the publisher aimed to assist other scholars and data scientists on Kaggle who are interested in creating models and conducting academic research. </p>This dataset includes a total of 1380 observations with six features. Several of them are __quantitative__ data such as ***earnings***, ***height***, ***ed(education level)***, and ***age***, while the others are __qualitative__ data such as ***sex*** and ***race***. Nonetheless, some elements of the quantitative data set were excluded from the dataset, such as the unit of the data. Also, the occupation of each data sample was not included in the dataset. For the data we used in this study, the race category consists of % white and % black, % hispanic and % of other races. Therefore, we expected to examine and gain insight from the wage gap between different gender, race and age groups. Each observation was composed of six attributes that demonstrated the potential wage gap between races, genders, and ages. Additionally, the publisher collected each sample's educational level and age besides race and earnings. Hence, the dataset not only represents people from different races but also could be categorized for different gender, age groups and education levels. As the dataset provides a wide range of sample's education levels and ages, we could better analyze each category and its relationships. </p>In terms of data validation, the dataset we used is held in a secure manner, as the dataset is from Kaggle, which datasets are posted and used for academic purposes only. The authenticity of Kaggle should be credited because Kaggle is an online website that provides a public data platform and a cloud-based workbench for data science. Hence, we obtained the dataset by downloading it from the website and uploading it to our repository. |
|5.0 Expected Implications | The expected implications relate to the biased job application screening processes which hinder the possibility of women of color getting hired. Such unfair processes used social identity and its proxies to exercise unfair employment practices that disregard the talents of minoritized groups. The practice of using proxies to race and gender such as the candidate's name, address, colleges, is used to filtered out the application of marginalized groups early on in the hiring process. It can be a window to look at the wage gap problem. Another implication can be the availability of job information which can be unfairly distributed, leading to the decreased exposure of minoritized groups to potential job openings. These implications urge policy makers and employers to be mindful of the current unjust system and their responsibility of making the necessary changes.|
|6.0 Limitations | Since what we are looking at are purely correlation, hence, they might no constitute strong arguments by themselves. What we are addressing are the very outer layers of the problem. Moreover, we believe there are other factors rooted from the systemic oppression that cause mediating effects, which then perpetuate the wage gap. The referred factors include, but not limit to, disparity in access to education, healthcare, green space, and job opportunity. Biases against certain social identities might also impact the likelihood of getting promoted, hence, perpetuate the existing wage gap. For this reason, we have to be cautious and mindful when using our editorial freedom in the data visualization. It is important to take these limitations into account when communicating to the target audiences through this dataset.|
|Acknowledgements | We would like to thank Professor David G Hendry and Teaching Assistant Thomas Wilson for supporting us throughout the project.|
|References | 1.  Chapman, Stephen J., and Nicole Benis. “Ceteris Non Paribus: The Intersectionality of Gender, Race, and Region in the Gender Wage Gap.” Women's Studies International Forum, vol. 65, Elsevier Ltd, 2017, pp. 78–86, https://doi.org/10.1016/j.wsif.2017.10.001. </p>  2. Misra, Joya, and Marta Murray-Close. “The Gender Wage Gap in the United States and Cross Nationally.” Sociology Compass, MisraJ., and Murray-CloseM. (2014), The Gender Wage Gap in the United States and Cross Nationally, Sociology Compass, 8, pages 1281-1295, doi: 10.1111/soc4.12213, vol. 8, no. 11, Blackwell Publishing Ltd, 2014, pp. 1281–95, https://doi.org/10.1111/soc4.12213.</p> 3.  Weinberger, Catherine J. “Race and Gender Wage Gaps in the Market for Recent College Graduates.” Industrial Relations (Berkeley), vol. 37, no. 1, Blackwell Publishers Inc, 1998, pp. 67–84, https://doi.org/10.1111/0019-8676.721998035. |

## ACCESSING THE /source DIRECTORY
```{r, echo = FALSE}
# Example 1: Note relative path, which can be read: Up one
# directory(..), down into source (/source), and 
# then "source" an R file (data_access.R)
source("../source/data_access.R")
source("../source/Graph_1.R")
source("../source/code_graph2.R")
source("../source/aggregate_table.R")
source("../source/summary.R")
source("../source/graph3.R")

data_access_test()
```

## Summary Tables
```{r summary_info, echo = FALSE}
str(summary_info)
```
The list above is contain important information regards to this data set. For example, number of trails, standard deviation of earning, max & min earning, and the average of earning. These information can help you understand the data set quickly.
## Aggregate Tables

* These aggregate table shows the average and median income between different groups of people on the basis of race and gender. The main focus of this project is to investigate the variability of income - an indicator of wealth, between different social groups determined by race and gender.

** Average income across racial groups and between genders

```{r aggregate_table, echo = FALSE}
library("knitr")
kable(aggregate_table)
```

** Median income across racial groups and between genders

```{r aggregate_table_2, echo = FALSE}
library("knitr")
kable(aggregate_table_2)
```


## Graphs


```{r grpah1, echo=FALSE}
plot(graph1)
```

The chart above indicates income distribution by race in white, black, Hispanic, and other groups, showing the potential earnings gaps between races. From the graph above, it is clear that most people earn between 0 and 50000 for all races, and the median income of white and other races is slightly higher than those of black and Hispanic communities. While most of the observation fall below 50000, all categories of observations appear outliners that may influence the chart, especially the white community, which includes many outliners that are exceptionally large. For instance, an observation from white races earns more than 30k which is much higher than other observations. These exceptional values affect the mean, pulling it to the right, resulting in a right-skewed distribution with a median greater than the mean.

```{r graph2, echo=FALSE}
plot(graph2)
```

This second graph shows the relationship between gender and the amount of wages earned. From this graph it can be seen that males earn a higher amount of money compared to females. The filled part of the graph shows the number of people earning a certain amount,looking at the distribution of male vs female the amount of wages earned seems more evenly distributed for males then it does for females.

**Median Earning among Different Races under 2 Genders**
```{r graph3, echo=FALSE}
plot(graph3)
```
```{r graph4, echo=FALSE}
plot(graph4)
```

Graph 3 is would be a combination of the above 3 graph, comparing the the different of the incomes between different genders and race. It is clear to observe that there is significant difference in earings among different ethnic groups. However, the ranking patterns in one group does not match between male and female. For example, white ethnic group is leading the earning in the male group, while it ranks the lowest in the female group