---
title: "Social Determinants of Wage Gaps"
author: "Claire Lai, Trang Tran, Tony Zhang, Jiali Liu"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## ACCESSING THE /source DIRECTORY
### Example 1: Source some code and run a function
```{r}
# Example 1: Note relative path, which can be read: Up one
# directory(..), down into source (/source), and 
# then "source" an R file (data_access.R)
source("../source/data_access.R")
source("../source/Graph_1.R")
source("../source/code_graph2.R")
source("../source/aggregate_table.R")
data_access_test()
```

## Aggregate Tables

* These aggregate table shows the average and median income between different groups of people on the basis of race and gender. The main focus of this project is to investigate the variability of income - an indicator of wealth, between different social groups determined by race and gender.

** Average income across racial groups and between genders

```{r aggregate_table, echo = FALSE}
library("knitr")
kable(aggregate_table)
```

** Median income across racial groups and between genders

```{r aggregate_table_2, echo = FALSE}
library("knitr")
kable(aggregate_table_2)
```


## Graphs


```{r grpah1, echo=FALSE}
plot(graph1)
```

The chart above indicates income distribution by race in white, black, Hispanic, and other groups, showing the potential earnings gaps between races. From the graph above, it is clear that most people earn between 0 and 50000 for all races, and the median income of white and other races is slightly higher than those of black and Hispanic communities. While most of the observation fall below 50000, all categories of observations appear outliners that may influence the chart, especially the white community, which includes many outliners that are exceptionally large. For instance, an observation from white races earns more than 30k which is much higher than other observations. These exceptional values affect the mean, pulling it to the right, resulting in a right-skewed distribution with a median greater than the mean.

```{r graph2, echo=FALSE}
plot(graph2)
```

This second graph shows the relationship between gender and the amount of wages earned. From this graph it can be seen that males earn a higher amount of money compared to females. The filled part of the graph shows the number of people earning a certain amount,looking at the distribution of male vs female the amount of wages earned seems more evenly distributed for males then it does for females.